---
title: "Take-home Exercise 1: Geospatial Analytics for Social Good"
editor: visual
---

## Background

In Singapore, people are used to access clean water any where and any time. The convenience of getting clean water makes some people forget how precious clean water is, can how difficult it is for people living in other countries to use clean water.

Nigeria is one of the countries still facing severe water crisis, and about one third of the children there don't have enough water to meet their daily needs.

In this take-home exercise, we are going to analyze the water point related data in Nigeria to reveal the spatial patterns of Not Functional water points.

## Problem Statement

-   Apply appropriate global and local measures of spatial association techniques on Nigeria data to study the spatial patterns of Not Functional water points.

## Getting Started

### Data sources

-   **Aspatial Data**

A set of a spatial data containing the water point related information is downloaded from [Water Point Data Exchange (WPdx)](https://data.waterpointdata.org/dataset/Water-Point-Data-Exchange-Plus-WPdx-/eqje-vguj/data) website.

The original data set contains the data from 17 countries, and only the subset of Nigeria is used for the analysis in this exercise.

**Geospatial Data**

Nigeria Level-2 Administrative Boundary polygon features GIS data downloaded from [geoBoundaries](https://www.geoboundaries.org/) will be used in this exercise.

### Loading R Packages

First of all, let's load the required R packages into R:

-   **sf**: importing, managing and processing geospatial data.

-   **spdep**: computing spatial weights and spatially lagged variables.

-   **tmap**: plotting Thematic Maps

-   **tidyverse**: importing, wrangling and visualizing data. It consists of a family of R packages, including **readr**, **readxl**, **tidyr**, **dplyr** and **ggplot2**.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, stringr)

```

### Importing Data

**Aspatial Data**

Let's import the aspatial data containing the water point information.

```{r}
#| eval: false
wp <- read_csv("Data/Water_Point_Data_Exchange-Plus_WPdx.csv")

```

The data set contains 406,566 rows and 70 columns.

```{r}
#| eval: false
table(wp$`#clean_country_name`)

#| eval: false
length(table(wp$`#clean_country_name`))

```

The aspatial data set contains data from 17 countries, and Nigeria consists of 95,008 rows.

Now, let's get the subset of Nigeria.

```{r}
#| eval: false
wp <- subset(wp, `#clean_country_name` == "Nigeria")

#| eval: false
list(wp)

```

The subset now contains only data for Nigeria, and it has 95,008 rows and 70 columns.

As our focus is to study the non-functional water points, we'll need to filter the data set where "#status" starts with "Non-functional".

```{r}
#| eval: false
# Extract the first word from #status column
wp$status_n <- word(wp$`#status`, 1)

#| eval: false
unique(wp$status_n)

```

```{r}
#| eval: false
# Filter the data set where status_n == Non-functional
wp <- subset(wp, status_n == "Non-functional")

```

The data set now contains only the data for non-functional water points, and it has 32,101 rows and 70 columns.

Now let's save the data set with the selected columns.

```{r}
#| eval: false
wp <- wp %>% select(c(`#lat_deg`,
                `#lon_deg`,
                `#water_source_category`,
                `#water_tech_category`,
                `#install_year`,
                `#management_clean`,
                `#subjective_quality`,
                water_point_population,
                local_population_1km,
                crucialness_score
))

```

```{r}
#| eval: false
write_csv(wp, "Data/Water_Point_Nigeria.csv")

```

After reducing the data file size, let's import the data again for analysis.

```{r}
wp <- read_csv("Data/Water_Point_Nigeria.csv")

```

The data set now contains 32,101 rows and the selected 10 columns.

## References

-   UNICEF. (2022, March 22). *Nearly one third of Nigerian children do not have enough water to meet their daily needs - UNICEF.* <https://www.unicef.org/nigeria/press-releases/nearly-one-third-nigerian-children-do-not-have-enough-water-meet-their-daily-needs>

#### 
