---
title: "In-class Exercise 1: Spatial Weights and Applications"
editor: visual
---

## Overview

In this in-class exercise, I learned how to compute spatial weights using R.

## Getting Started

Firstly, we need to install the required R packages.

-   **sf**: importing, managing and processing geospatial data.

-   **spdep**: computing spatial weights and spatially lagged variables.

-   **tmap**: plotting Thematic Maps.

-   **tidyverse**: importing, wrangling and visualizing data. It consists of a family of R packages, including **readr**, **readxl**, **tidyr**, **dplyr** and **ggplot2**.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)

```

## Importing Required Data

### Importing polygon feature data

The following codes import *Hunan* shapfile into R as a polygon feature data frame.

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")

```

### Importing the aspatial data

The following code chunk imports *Hunan_2012* data set into R as a dataframe.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")

```

### Performing relational join

The code chunk below joins the two dataframes, adding the attributes in *hunan2012* dataframes to *hunan* dataframe.

```{r}
hunan <- left_join(hunan, hunan2012)

```

## Visualizing Weights and Applications

### Visualizing regional development indicator

The code chunk below plots two maps, one base map showing the county boundaries within Hunan province, and the other one is a choropleth map showing the distribution of GDPPC 2012 of the counties.

```{r}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size = 0.5)

gdppc <- qtm(hunan, "GDPPC")

tmap_arrange(basemap, gdppc, asp = 1, ncol = 2)

```

## Computing Contiguity Spatial Weights

### Computing (Queen) contiguity based neighbors

The code chunk below computes Queen contiguity weight matrix on *hunan* dataframe.

```{r}
wm_q <- poly2nb(hunan, queen = T)
summary(wm_q)

```

Check the neighbors for the 1st polygon in the object.

```{r}
wm_q[[1]]

```

Get the county name of polygon ID = 1.

```{r}
hunan$County[1]

```

Get the county names of the 5 neighbors of the 1st polygon in the object.

```{r}
hunan$NAME_3[wm_q[[1]]]

```

Get the GDPPC of these 5 neighbors.

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1

```

Display the complete weight matrix.

```{r}
str(wm_q)

```

### Creating (Rook) contiguity based neighbors

The code chunk below computes Rook contiguity weight matrix.

```{r}
wm_r <- poly2nb(hunan, queen = F)
summary(wm_r)

```

### Visualizing contiguity weights

Get the longitude of the counties from *hunan* dataframe.

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])

```

Get the latitude of the counties from *hunan* dataframe.

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])

```

Combine *longitude* and *latitude* into one object.

```{r}
coords <- cbind(longitude, latitude)

```

Check a few rows to make sure the data is combined correctly.

```{r}
head(coords)

```

#### Plotting Queen contiguity based neighbors map

```{r}
plot(hunan$geometry, border = "lightgray")
plot(wm_q, coords, pch = 19, cex = 0.6, add = T, col = "red")

```

#### Plotting Rook contiguity based neighbors map

```{r}
plot(hunan$geometry, border = "lightgray")
plot(wm_r, coords, pch = 19, cex = 0.6, add = T, col = "red")

```

#### Plotting Both Queen and Rook contiguity based neighbors maps

```{r}
par(mfrow = c(1,2))

plot(hunan$geometry, border = "lightgray")
plot(wm_q, coords, pch = 19, cex = 0.6, add = T, col = "red")

plot(hunan$geometry, border = "lightgray")
plot(wm_r, coords, pch = 19, cex = 0.6, add = T, col = "red")

```

## Computing Distanced Based Neighbors

### Determining the cut-off distance

The code chunk below calculates and displays the distribution of distance among all neighbors.

```{r}
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = T))
summary(k1dists)

```

The maximum distance could be used as the upper threshold to every county has at least one neighbor.

### Computing fixed distance weight matrix

The code chunk below calculates the distance weight matrix.

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = T)
wm_d62

```

**Quiz**: What is the meaning of "Average number of links: 3.681818" shown above?

**Ans**: Average number of links above shows that each county has 3.68 neighbors within the threshold on average.

Display the content of wm_d62 weight matrix.

```{r}
str(wm_d62)

```

Another way to display the weight matrix.

```{r}
table(hunan$County, card(wm_d62))

```

Compute the number of disjoint connected counties.

```{r}
n_comp <- n.comp.nb(wm_d62)
n_comp$nc

```

Get the id of the disjoint connected counties.

```{r}
table(n_comp$comp.id)

```

#### Plotting dixed distance weight matrix

```{r}
plot(hunan$geometry, border = "lightgray")
plot(wm_d62, coords, add = T)
plot(k1, coords, add = T, col = "red", length = 0.08)

```

Plot the two graphs separately.

```{r}
par(mfrow = c(1,2))

plot(hunan$geometry, border = "lightgray")
plot(k1, coords, add = T, col = "red", length = 0.08, main = "1st nearest neighbours")

plot(hunan$geometry, border = "lightgray")
plot(wm_d62, coords, add = T, pch = 19, cex = 0.6, main = "Distance link")

```

### Computing adaptive distance weight matrix

The code chunk below controls the number of neighbors directly using k-nearest neighbors, either accepting asymmetric neighbors or imposing symmetry.

```{r}
knn6 <- knn2nb(knearneigh(coords, k = 6))
knn6

```

Display the content of the matrix.

```{r}
str(knn6)

```

#### Plotting distance based neighbors

```{r}
plot(hunan$geometry, border = "lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = T, col = "red")

```

## Weights Based on IDW

The code chunk below derives a spatial weight matrix based on Inversed Distance method.

```{r}
dist <- nbdists(wm_q, coords, longlat = T)
ids <- lapply(dist, function(x) 1/(x))
ids

```

### Row-standardized weights matrix

The code chunk below standardized the weights matrix by row. Style = "W" assigns equal weight to each neighboring polygon.

```{r}
rswm_q <- nb2listw(wm_q, style="W", zero.policy = T)
rswm_q

```

Check the weight of the 1st polygon's neighbors.

```{r}
rswm_q$weights[10]

```

Derive a row standardized distance weight matrix

```{r}
rswm_ids <- nb2listw(wm_q, glist=ids, style="B", zero.policy=T)
rswm_ids

```

Display the weights of the neighbors of 1st polygon.

```{r}
rswm_ids$weights[1]

```

Check the distribution of the weights.

```{r}
summary(unlist(rswm_ids$weights))

```

## Application of Spatial Weight Matrix

### Spatial lag with row-standardized weights

The code chunk below computes the average neighbor GDPPC value for each polygon, which is also known as **spatially lagged values**.

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag

```

Append the spatially lag GDPPC values to *hunan* dataframe.

```{r}
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan,lag.res)

```

Display the the average neighboring income values for each county.

```{r}
head(hunan)

```

Plot both the GDPPC and spatial lag GDPPC for comparison.

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp = 1, ncol = 2)

```

### Spatial lag as a sum of neighboring values

The code chunk below calculates the spatial lag as a sum of neighboring values by assigning binary weights.

```{r}
b_weights <- lapply(wm_q, function(x) 0 * x + 1)
b_weights2 <- nb2listw(wm_q, 
                       glist = b_weights, 
                       style = "B")
b_weights2

```

Compute a lag variable from the weight and GDPPC.

```{r}
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")

```

Take a look at the results.

```{r}
lag_sum

```

Append the *lag_sum* attribute into *hunan* dataframe.

```{r}
hunan <- left_join(hunan, lag.res)

```

Plot both the *GDPPC* and *Spatial Lag Sum GDPPC* for comparison.

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp = 1, ncol = 2)

```

### Spatial window average

The spatial window average uses row-standardized weights and includes the diagonal element.

```{r}
wm_q1 <- wm_q

```

Add the diagonal element to the neighbour list.

```{r}
include.self(wm_q1)

```

```{r}
wm_q1 <- nb2listw(wm_q1)
wm_q1

```

Create the lag variable from our weight structure and GDPPC variable.

```{r}
lag_w_avg_gpdpc <- lag.listw(wm_q1, 
                             hunan$GDPPC)
lag_w_avg_gpdpc

```

Convert the lag variable *listw* object into a dataframe.

```{r}
lag.list.wm_q1 <- list(hunan$NAME_3, lag.listw(wm_q1, hunan$GDPPC))
lag_wm_q1.res <- as.data.frame(lag.list.wm_q1)
colnames(lag_wm_q1.res) <- c("NAME_3", "lag_window_avg GDPPC")

```

Append *lag_window_avg GDPPC* values onto *hunan* dataframe.

```{r}
hunan <- left_join(hunan, lag_wm_q1.res)

```

Plot the *GDPPC* and *lag_window_avg* GDPPC map next to each other for quick comparison.

```{r}
gdppc <- qtm(hunan, "GDPPC")
w_avg_gdppc <- qtm(hunan, "lag_window_avg GDPPC")
tmap_arrange(gdppc, w_avg_gdppc, asp=1, ncol=2)

```

### Spatial window sum

The spatial window sum is the counter part of the window average, but without using row-standardized weights.

```{r}
wm_q1 <- wm_q

```

Add the diagonal element to the neighbor list.

```{r}
include.self(wm_q1)

```

```{r}
wm_q1

```

Assign binary weights to the neighbor structure that includes the diagonal element.

```{r}
b_weights <- lapply(wm_q1, function(x) 0 * x + 1)
b_weights[1]

```

Assign weight values.

```{r}
b_weights2 <- nb2listw(wm_q1, 
                       glist = b_weights, 
                       style = "B")
b_weights2

```

Compute the lag variable.

```{r}
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc

```

Convert the lag variable *listw* object into a dataframe.

```{r}
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")

```

Append *w_sum GDPPC* values to *hunan* dataframe.

```{r}
hunan <- left_join(hunan, w_sum_gdppc.res)

```

Plot the *GDPPC* and *lag_sum* GDPPC map next to each other for quick comparison.

```{r}
gdppc <- qtm(hunan, "GDPPC")
w_sum_gdppc <- qtm(hunan, "w_sum GDPPC")
tmap_arrange(gdppc, w_sum_gdppc, asp = 1, ncol = 2)

```
